---
title: "R Packages for Advanced Panel Data Econometrics"
abstract: >
  We present seven R packages addressing gaps in panel data econometrics. The
  xtmispanel package provides missing data diagnostics and imputation for panel
  data. The xtcsdq package tests for cross-sectional dependence in panel quantile
  regressions following Demetrescu, Hosseinkouchack, and Rodrigues (2023). The
  xtpcaus package implements panel Fourier Toda--Yamamoto and panel quantile
  causality tests. The xtpcmg package provides group-mean and pooled FM-OLS for
  panel cointegrating polynomial regressions. The xtrec package implements
  recursively detrended panel unit root tests of Westerlund (2015). The xtfifevd
  package estimates time-invariant regressor effects in fixed effects models. The
  xtqsh package provides quantile regression slope homogeneity tests following
  Galvao, Juhl, Montes-Rojas, and Olmo (2017). These packages extend the panel
  data toolkit beyond plm and fixest, covering missing data, quantile methods,
  polynomial cointegration, and advanced diagnostics. All packages are
  open-source under the GPL-3 license.
author:
  - name: Muhammad Abdullah Alkhalaf
    affiliation: Rufyq Elngeh for Academic and Business Services
    address: Riyadh, Saudi Arabia
    orcid: 0009-0002-2677-9246
    email: muhammedalkhalaf@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: paper.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, comment = "#>")
```

# Introduction

R provides foundational panel data tools through the `plm` package
\citep{Croissant2008} and the `fixest` package \citep{Berge2021}. However,
several recent methodological advances lack R implementations: specialized panel
missing data diagnostics, cross-sectional dependence tests for quantile
regressions, panel Fourier Granger causality, cointegrating polynomial
regressions with FM-OLS, recursively detrended unit root tests, estimation of
time-invariant regressor effects, and quantile slope homogeneity tests.

This article introduces seven R packages addressing these gaps: `xtmispanel`,
`xtcsdq`, `xtpcaus`, `xtpcmg`, `xtrec`, `xtfifevd`, and `xtqsh`.

# The xtmispanel package

The `xtmispanel` package detects, diagnoses, and imputes missing values in panel
data. It provides per-variable, per-panel, and per-period summaries, approximate
\citet{Little1988} MCAR and MAR tests, and eleven imputation methods.

```{r xtmispanel-example}
library(xtmispanel)

# Create panel data with missing values
set.seed(42)
N <- 20; T_obs <- 30
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  gdp = rnorm(N * T_obs, 100, 15),
  investment = rnorm(N * T_obs, 20, 5)
)
# Introduce 10% missing values
miss_idx <- sample(nrow(panel_data), size = 0.1 * nrow(panel_data))
panel_data$gdp[miss_idx] <- NA

# Detect, test, and impute
result <- xtmispanel(panel_data, index = c("country", "year"),
                     detect = TRUE, test = TRUE,
                     impute = "pmm", target = "gdp")
```

The available imputation methods include mean, median, LOCF, NOCB, linear and
spline interpolation, predictive mean matching (PMM), hot-deck, KNN, random
forest, and EM algorithm.

# The xtcsdq package

The `xtcsdq` package tests for cross-sectional error dependence in panel quantile
regressions following \citet{Demetrescu2023}. It provides T$_\tau$ and
bias-corrected $\tilde{T}_\tau$ statistics, plus a portmanteau M$_K$ statistic
across quantiles.

```{r xtcsdq-example}
library(xtcsdq)

set.seed(123)
N <- 15; T_obs <- 40
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  y = rnorm(N * T_obs),
  x1 = rnorm(N * T_obs),
  x2 = rnorm(N * T_obs)
)

result <- xtcsdq(y ~ x1 + x2, data = panel_data,
                 index = c("country", "year"),
                 quantiles = c(0.25, 0.5, 0.75))
summary(result)
```

# The xtpcaus package

The `xtpcaus` package implements Panel Fourier Toda--Yamamoto (PFTY) and Panel
Quantile Causality (PQC) tests with bootstrap p-values.

```{r xtpcaus-example}
library(xtpcaus)

set.seed(456)
N <- 10; T_obs <- 50
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  gdp = unlist(lapply(1:N, function(i) cumsum(rnorm(T_obs, 0.02, 0.1)))),
  investment = unlist(lapply(1:N, function(i) cumsum(rnorm(T_obs, 0.01, 0.08))))
)

result <- xtpcaus(data = panel_data, y = "gdp", x = "investment",
                  panel_id = "country", time_id = "year",
                  test = "pfty", kmax = 3, nboot = 499)
summary(result)
```

# The xtpcmg package

The `xtpcmg` package provides Group-Mean and Pooled FM-OLS for panel
cointegrating polynomial regressions. Group-Mean follows \citet{Wagner2023};
Pooled follows \citet{deJong2022}. It supports quadratic and cubic polynomials
with HAC long-run variance, turning point analysis, and the \citet{Swamy1970}
slope homogeneity test.

```{r xtpcmg-example}
library(xtpcmg)

set.seed(789)
N <- 15; T_obs <- 60
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  emissions = unlist(lapply(1:N, function(i) {
    gdp <- cumsum(rnorm(T_obs, 0.05, 0.1))
    0.5 * gdp - 0.02 * gdp^2 + rnorm(T_obs, 0, 0.3)
  })),
  gdp = unlist(lapply(1:N, function(i) cumsum(rnorm(T_obs, 0.05, 0.1))))
)

# Group-Mean FM-OLS with quadratic polynomial
result <- xtpcmg(data = panel_data, y = "emissions",
                 x = "gdp", panel_id = "country",
                 time_id = "year", model = "mg", q = 2)
summary(result)
```

This package is particularly useful for testing the Environmental Kuznets Curve
hypothesis in a panel data setting.

# The xtrec package

The `xtrec` package implements the recursively detrended panel unit root tests of
\citet{Westerlund2015}. The basic t-REC assumes iid errors; the robust t-RREC
accounts for serial correlation, cross-sectional dependence, and
heteroskedasticity.

```{r xtrec-example}
library(xtrec)

set.seed(101)
N <- 20; T_obs <- 50
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  gdp = unlist(lapply(1:N, function(i) cumsum(rnorm(T_obs))))
)

result <- xtrec(data = panel_data, var = "gdp",
                panel_id = "country", time_id = "year",
                robust = TRUE)
summary(result)
```

# The xtfifevd package

The `xtfifevd` package implements FEVD \citep{Plumper2007}, FEF, and FEF-IV
\citep{PesaranZhou2016} estimators for time-invariant regressors in panel fixed
effects models.

```{r xtfifevd-example}
library(xtfifevd)

set.seed(202)
N <- 25; T_obs <- 20
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  y = rnorm(N * T_obs),
  x1 = rnorm(N * T_obs),  # time-varying
  x2 = rnorm(N * T_obs),  # time-varying
  z1 = rep(rnorm(N), each = T_obs),  # time-invariant
  z2 = rep(rnorm(N), each = T_obs)   # time-invariant
)

result <- xtfifevd(y ~ x1 + x2 | z1 + z2, data = panel_data,
                   id = "country", time = "year",
                   method = "fef")
summary(result)
```

# The xtqsh package

The `xtqsh` package implements the quantile regression slope homogeneity test of
\citet{Galvao2017}. It provides $\hat{S}$ (chi-squared) and $\hat{D}$ (standard
normal) statistics for joint and marginal homogeneity.

```{r xtqsh-example}
library(xtqsh)

set.seed(303)
N <- 20; T_obs <- 30
panel_data <- data.frame(
  country = rep(paste0("C", 1:N), each = T_obs),
  year = rep(2001:(2000 + T_obs), N),
  y = rnorm(N * T_obs),
  x1 = rnorm(N * T_obs),
  x2 = rnorm(N * T_obs)
)

result <- xtqsh(y ~ x1 + x2, data = panel_data,
                id = "country", time = "year",
                tau = c(0.25, 0.5, 0.75))
summary(result)
```

# Comparison with existing implementations

The `plm` package \citep{Croissant2008} provides standard panel data estimators
(pooled OLS, fixed effects, random effects, first differences) and several
diagnostic tests (Hausman, serial correlation, cross-sectional dependence). The
`fixest` package \citep{Berge2021} offers fast multi-way fixed effects estimation.
However, neither covers:

- **Missing data diagnostics** specific to panel structure (provided by `xtmispanel`)
- **Cross-sectional dependence tests for quantile regressions** (provided by `xtcsdq`)
- **Panel Fourier Granger causality** (provided by `xtpcaus`)
- **Cointegrating polynomial regressions** with FM-OLS (provided by `xtpcmg`)
- **Recursively detrended panel unit root tests** (provided by `xtrec`)
- **Time-invariant regressor estimation** beyond basic Hausman--Taylor (provided by `xtfifevd`)
- **Quantile slope homogeneity tests** (provided by `xtqsh`)

The `panelvar` package provides panel VAR estimation but not the Fourier and
quantile causality tests in `xtpcaus`. The `cointReg` package offers FM-OLS for
cointegrating regressions but not the polynomial panel extension in `xtpcmg`.

# Summary

The seven packages presented here extend R's panel data toolkit in important
directions: missing data management, quantile methods, polynomial cointegration,
recursive detrending, time-invariant regressors, and advanced diagnostics. All
packages are available on CRAN under GPL-3.

\bibliography{paper}
